Stability in human interaction networks: primitive typology of vertex, prominence of measures and activity statistics

### 2015-01-20 paper.tex
revisado todo artigo, menos tabelas, figuras e referencias

\usepackage{graphicx}% Include figure files
\usepackage{grffile}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\usepackage{multirow}
\usepackage{color} % for the notes
\usepackage{etex}
\reserveinserts{58}
\usepackage{morefloats}
\usepackage{hyperref}

@@

\begin{abstract}
    This article reports a characterization of interaction networks and its stability. Such a task involves a selection of aspects to investigate, which lead to: 1) activity distribution in time and among participants, 2) a sound classification of vertex: peripheral, intermediary and hub sectors, 3) combination of basic measures into components with greater dispersion. While time patterns of activity are not obvious, participant activity follows a scale-free distribution. Comparison with ideal Erd\"os-R\'enyi network with the same number of edges and vertexes was a sound criterion  for distinguishing sectors on the networks. Principal components in basic measures spaces revealed interesting and regular patterns of independence and dispersion. This includes a ranking of measures that most contribute to dispersion: 1) degree and strength measures, 2) symmetry related quantization, and 3) clusterization. Results suggest typologies for these networks and participants. Further work include considerations of text production, psychoanalysis inspired typologies, participatory democracy exploitation of observed properties, and better visualization support for network evolution.
\end{abstract}

@@

While significant measures will depend on the model and system characteristics~\cite{newmanStru, newmanWeight},
this work considers only directed, weighted and human interaction networks. Undirected and unweighted representation of such networks is also seen in the literature and can be obtained by simplification~\cite{GMANE2}.

Text mining and typologies of online participants benefit from results here presented~\cite{rcText,rcTipo}. 
Although all networks considered originated from email lists,
 coherence with literature suggests that results hold for a more general class of interaction networks,
 such as observed in online platforms (LinkedIn, Facebook, Twitter).

\subsection{Related work}
 Works on network evolution often consider solely network growth, in which there is a monotonic increase in the number of events considered~\cite{barabasiEvo}. Moreover, selected exceptions are reported in this section.
The evolution considered in the present article is characterized by a constant number of messages, which is also present in literature, but was less explored to date.

 The evolution of interaction networks was addressed with focus on community evolution, a work that ignores the direction of edges~\cite{barabasiEvo}. Two topologically different networks are reported to emerge, depending on the frequency of interactions, which present a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. 
Recent work on email lists consider an isolated snapshot in order to verify or draw hypothesis. In such, free-scale properties were verified~\cite{bird}, and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

 Such results are in accordance with phenomena observed in this work and linguistic characterization is being described in a deriving article~\cite{rcText}. See Appendix~\ref{sec:fure} for further considerations of related work.

\section{Data description}
    \subsection{Email lists and messages}
Email list messages were obtained from
the GMANE email archive~\cite{GMANE}, which consists of more than 20,000 email lists and more than 130,000,000 messages~\cite{GMANEwikipedia}. These lists cover a variety of topics, mostly technology-related. It can be described as a corpus with metadata of its messages, like send time, place, sender name, sender email address etc.
GMANE usage in scientific research is reported in studies of isolated lists and of lexical innovations~\cite{GMANE2,bird}. Appendix~\ref{scripts} is dedicated scripts for gathering and processing GMANE email messages.

    \subsection{Chosen dataset}
The four lists below were selected for their diversity, easing initial observance of natural and general properties.
\begin{itemize}
    \item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE archive.}. Participants are from different countries, and English is the language used the most. More technical and less active version of LAU. Abbreviated LAD from now on.
    \item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE archive.}. Dominated by specialized computer programmers. Participants are from different countries, and English is the language used the most. Abbreviated as CPP from now on.
    \item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE archive.}. Dominated by Brazilian activists and digital culture interests. Participants are mostly Brazilians, and Portuguese is the most used language, although Spanish and English usage is commonplace. Abbreviated MET from now on.
    \item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE archive.}. Dominated by participants with hybrid artistic and technological interests. Participants are from different countries, and English is the language used the most. Abbreviated as LAU from now on.
\end{itemize}

@@

\section{Characterization methods}
After immersion in the data and on appropriated literature, the following methods for network characterization were chosen: 1) statistics of activity along time; 2) division of network in hubs, intermediary and peripheral vertex; 3) prominence of topological measures; 4) evolutive visualizations and quantitative observations; 5) typological elaborations of networks and participants.
Each of these methods are described bellow, as other handy alternatives.

    \subsection{Temporal activity statistics}
Number of messages along time with respect to seconds, minutes, hours, days of the week or the month, and months of the year. These were displayed as tables in Appendix~\ref{tabTime}, results are outlined in Section~\ref{constDisc}.

    \subsection{Interaction network}\label{intNet}
Regarding literature~\cite{bird,newmanCommunityDirected,newmanCommunity2013}, interaction networks might be both weighted or unweighted, both directed or undirected. Networks in this article are directed and weighted, considered as more informative among possibilities (directed unweighted, and undirected weighted and undirected unweighted).
More precisely, the networks reported are erected as follows: a direct response from participant B to participant A forms an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to A, he read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network, as B read the message and considered what A wrote worth of responding, giving status to A, thus $B\rightarrow A$. This article uses the information network described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to edge weight. Self-loops were regarded as non-informative and discarded. This networks are described as exhibiting free-scale and small world properties, as expected for a social network~\cite{bird}.

@@

Previous messages on the thread create directed edges from their author to the observed message's author. Edges can be created from all antecedent messages on the message-response thread. Is this work, only immediate predecessors are linked to new message's author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a connection between $x\rightarrow z$. Potential interpretations for this weaker connection are usually common sense, such as: double length, half weight or with one more ``obstacle''. This suggests the adoption of other centrality measures that account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

@@

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser}
    \caption{Degree distribution on scale-free and Erd\"os-R\'enyi ideal networks. The later has more
        intermediary, as the former has more peripheral and hubs. Sections are
        given by the two intersections of the distributions $k_1$ and $k_2$. Characteristic degrees
    are in compact intervals of degree: $[0,k_1]$, $(k_1,k_2]$, $(k_2,k_{max}]$ for the three sections considered (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

Because of social networks' tendency to have scale-free distribution of properties,
one can compare it to an Erd\"os-R\'enyi random graph and consider peripheral, intermediary and hub
sectors~\cite{3setores}, as depicted in Figure~\ref{fig:setores}.
The degree distribution $\widetilde{P}(k)$ of an ideal
scale-free network $\mathcal{N}_f$ with $N$ vertexes and $z$ edges, has less
average degree nodes when compared with the distribution $P(k)$ of an Erd\"os-R\'enyi
random graph with the same number of vertexes and edges:

@@

If $\mathcal{N}_f$ is directed and has self-loops, the probability
of the presence of an unknown edge is $p=\frac{z}{N(N-1)}$, where $N(N-1)$ is the maximum number of edges for a network with $N$ vertexes, directed edges and without selfloops.
A vertex in the ideal Erd\"os-R\'enyi digraph with the same number of vertexes and edges, and thus the same probability $p$ for the presence of an edge, will have degree $k$ with probability:

@@

The arguments behind this classification are: 1) vertexes so connected that they are virtually inexistent in networks connected at pure chance, specially without preferential attachment, are correctly associated to hubs sector. Vertexes with very few connections, which are way more abundant than expected by pure chance, are correctly associated to periphery. Degree values predicted as the most abundant if connections are created by pure chance, near the average, and less frequent in free-scale phenomena, are correctly associated to intermediary vertexes.

To assure statistical validity, bins can be chosen to span the average of $\eta$ gaps between measure values. Thus, each bin, starting at degree $k_i$, spans $\Delta_i=\frac{\sum_{i=1}^{\eta}(k_{i+1}-k_i)}{\eta}$ values, with borders $k_i$ and $k_i+\Delta_{i}$. This changes equation~\ref{criterio} to:

\begin{equation}\label{criterio2}
    \sum_{x=k_i}^{k_i+\Delta_i} \widetilde{P}(x) < \sum_{x=k_i}^{k_i+\Delta_i} P(x) \Rightarrow \text{i is intermediary}
\end{equation}

If instead strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}$ the average weight of an edge and $s_i$ the vertex strength. For in and out degrees and strengths, comparisons should be made with $\kappa_i=2k_i^{in}$, $\kappa_i=2k_i^{out}$, $\kappa_i=2\frac{s_i^{in}}{\overline{w}}$ and $\kappa_i=2\frac{s_i^{out}}{\overline{w}}$. Results of these segmentations are discussed in subsection~\ref{subsec:pih}.

@@

\begin{itemize}
    \item Exclusivist criteria: vertex are only classified if the class is the same with respect to all measures. In this case, total vertex classified (usually) does not reach 100\%, which is specified by a black line in Appendix~\ref{figures}.
    \item Inclusivist criteria: vertex has class given by any of the measures. In this case, a vertex can have more than one class and the fraction of class attribution beyond total number of vertexes is also represented by a black line in Appendix~\ref{figures}.
    \item Exclusivist cascade: hubs are only hubs if classified as hub with respect to all measures. Intermediary are the vertexes classified as intermediary or hub with respect to all measures. Vertexes left are regarded as peripheral vertex.
    \item Inclusivist cascade: hubs are vertexes classified as hubs by any of the measures. From the vertexes left, if any is classified as intermediary by any measure, than it is intermediary. The rest of the vertexes are peripheral.
    \item Exclusivist externals: hubs have unanimous classification with respect to all measures. Of vertexes left, peripheral vertexes are the ones classified as hub or peripheral by simple criterion. The rest represent intermediary sector.
    \item Inclusivist externals: hubs are vertexes classified as hubs with respect to any measure. From vertexes left, if a vertex is classified as peripheral with respect to any measure, than it is peripheral. The rest is regarded intermediary sector.
\end{itemize}

@@

        \subsubsection{Topological measures}\label{measures}
This article
restricts this analysis to a small selection of the most basic
measures of each vertex:

\begin{itemize}
    \item Degree $d_i$: number edges linked to node $i$.
    \item In-degree $d_i^{in}$: number of edges ending at node $i$.
    \item Out-degree $d_i^{out}$: number of edges departing from node $i$.
    \item Strength $s$: sum of weights of all edges linked to node $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at node $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from node $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked. Standard clustering coefficient for undirected graphs was used.
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain the node $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

@@

\begin{itemize}
    \item asymmetry of note $i$: $asy_i=\frac{d_i^{in}-d_i^{out}}{d_i}$.
    \item mean of asymmetry of edges: $\mu_i^{asy}=\frac{\sum_{j\in J_i} e_{ji}-e_{ij}}{|J_i|}$. Where $e_{xy}$ is 1 if there is and edge from $x$ to $y$, $0$ otherwise. $|J_i|$ is the number of neighbors of vertex $i$.
    \item standard deviation of asymmetry of edges: $\sigma_i^{asy}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{asy} -(e_{ji}-e_{ij}) ]^2  }{|J_i|}  }$
    \item disequilibrium: $dis_i=\frac{s_i^{in}-s_i^{out}}{s_i}$.
    \item mean of disequilibrium of edges: $\mu_i^{dis}=\frac{\sum_{j \in J_i}\frac{w_{ji}-w_{ij}}{s_i}}{|J_i|}$, where $w_{xy}$ is weight of edge $x\rightarrow y$ and zero if there is no such edge.
    \item standard deviation of disequilibrium of edges: $\sigma_i^{dis}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{dis}-\frac{(w_{ji}-w_{ij})}{s_i}]^2}{|J_i|}}$
\end{itemize}

   \subsection{Evolutive observations}
Evolution of network is observed within a fixed number of messages (window size: $ws$) that shifts in the message timeline.
 All $ws=$50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000 were used. Within a same $ws$, the number vertex and edges vary in time, as do other network characteristics. Further work should deepen inspection of measure interdependence, this article holds to measures in Section~\ref{measures}.

        \subsubsection*{Visualization of network evolution}
In refining hypotheses, visualization of the network was crucial. Animations, image galleries and an online gadgets were made~\cite{animacoes,galGMANE,appGMANE}. Mapping of various topological measures to glyphs and layouts are being further explored as a parallel research. Furthermore, stable aspects of measures prominence along time are captured through mean and standard deviation (see Section~\ref{measures}). Constant sector sizes along time are observed in a timeline fashion in Appendix~\ref{figures}.

@@

        \subsubsection{Hours of the day}
Table~\ref{dia} shows how the four lists distribute activity along the day.
Afternoon was the most active period 6h of the day. Second 12h more active than first 12h. Even so, activity peak occurs around midday, with a slight skew towards earlier hours.

        \subsubsection{Days of the week}
 Weekdays also exhibit an interesting pattern, to which is dedicated Table~\ref{semana}.
Days exhibit pattern on weekdays, with a decrease of at least one third and reaching two thirds on weekends.
        \subsubsection{Days along the month}
 Table~\ref{mes} shows activity along the month.

        \subsubsection{Months and larger divisions of the year}
Table~\ref{ano} is dedicated to activity in months and larger divisions of the year.

As for months along the year, there seems to be two periods of more prominent activities: Jun-Aug (MET and LAD), and Dec-Mar (CPP, LAU and LAD). These observations fit academic calendars, vacations and end-of-year holidays.

 Variation of activity in the days along the month are less prominent, one cannot point much more than a (probably not statistically relevant) tendency to first and second weeks to be more active. The most important result in the days along the month is their homogeneity with respect to activity. Last days of the month (29, 30 and 31) are not present in every month, and observed activity is proportional to incidence rates.

@@

As specified in Section~\ref{sectioning}, in order to classify nodes as hubs, intermediary or peripheral, $\widetilde{P}(k)$ is compared
to $P(k)$. For the networks studied here, i.e. derived from public email lists, if degree distribution is used classification, hubs sector size reaches peaks with $10\%$ of all vertexes. 
 If strength is used for comparison ($\widetilde{P}(k_i=\frac{s_i}{\overline{w}})$) is compared
to $P(k)$), hubs account for approximately $5\%$ of all vertex, i.e. strength classification yields half the number of hubs as plain degree. 
This results hold for in and out degrees and strengths,

Classification criteria exposed in subsection~\ref{sectioning} was used efficiently with windows with at least 200 messages. Specially with 1000 or more messages, criteria yields stable fractions of $\approx 5\%$ of hubs, $\approx [15-20]\%$ of intermediary and $\approx [75-80]\%$ peripheral vertex. The compound criteria, also described in Section~\ref{sectioning}, can be used as a classification refinement. This is specially useful in dealing with fewer messages, in which case the structure degenerates with respect to some of the degree and strength measures, but not all.

For the networks analysed, differences of using this smoothing process were not significant. There were between 20 and 1200 participants, so each participant were between $5\%$ and $0.08\%$ of all participants.  The bottom line is: for network sizes considered, if connectivity intensity would most probably not exist in an Ed\"os Renyi network, than it is not an intermediary intensity. As peripheral vertex are abundant, this statistic discussion has no relevance.

A reasonable window size for observation can be inferred by monitoring the giant component size and the degeneration of the hub, intermediary and peripheral sections. This degeneration is critical in the span of 50-100 messages. With compound criteria, such as exclusive cascade of Figure~\ref{fig:cpp250_}, the network seems to hold basic structure even with as few as 20-50 messages. This indicates that concentration of activity and low-activity participants occurs even with very few messages.

@@

This work aimed at finding common characteristics among (email) interaction networks. The analysis  involved primary measures observance and a formal criteria for coherent ratios of hub, intermediary and hub sectors. Nevertheless, inspection done by visualizations and raw data manipulations suggest peculiarities of interest, specially:

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{grffile}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\usepackage{multirow}
\usepackage{color} % for the notes
\usepackage{etex}
\reserveinserts{58}
\usepackage{morefloats}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
        colorlinks,
        linkcolor={red!50!black},
        citecolor={blue!50!black},
        urlcolor={blue!80!black}
}

@@

\begin{abstract}
    This article reports a characterization of interaction networks and its stability. Such a task involves a selection of aspects to investigate, which lead to: 1) activity distribution in time and among participants; 2) a sound classification of vertex: peripheral, intermediary and hub sectors; 3) combination of basic measures into components with greater dispersion (PCA). While time patterns of activity are not obvious, participant activity follows a scale-free distribution. Comparison of incident networks with ideal Erd\"os-R\'enyi networks bearing the same number of edges and vertexes yields a sound criterion for distinguishing sectors on the networks. Principal components in basic measures space revealed interesting and regular patterns of independence and dispersion. This includes a ranking of measures that most contribute to dispersion: 1) degree and strength measures, 2) symmetry related quantization, and 3) clusterization. Results suggest typologies for these networks and participants. Further work include considerations of text production, psychoanalysis inspired typologies, participatory democracy exploitation of observed properties, and better visualization support for network evolution.
\end{abstract}

@@

While significant measures will depend on the model and system characteristics~\cite{newmanStru, newmanWeight},
this work considers only directed, weighted and human interaction networks. Undirected and unweighted representation of such networks is also found in the literature and can be obtained by simplification~\cite{GMANE2}.

Text mining and typologies of online participants benefit from the results here presented~\cite{rcText,rcTipo}. 
Although all networks considered originated from email lists,
 coherence with literature suggests that results hold for a more general class of interaction networks,
 such as observed in online platforms (e.g. LinkedIn, Facebook, Twitter).

\subsection{Related work}
 Works on network evolution often consider solely network growth, in which there is a monotonic increase in the number of events considered~\cite{barabasiEvo}. Moreover, selected exceptions are reported in this section.
The evolution considered in the present article is characterized by a constant number of messages, which is also present in literature, but less explored to date~\cite{netWindowEvo}.

 The evolution of interaction networks was addressed with a community focus, in a work that ignores the direction of edges~\cite{barabasiEvo}. Two topologically different networks are reported to emerge, depending on the frequency of interactions, which present a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. 
Recent work on email lists consider an isolated snapshot in order to verify or draw hypothesis. In such, free-scale properties were verified~\cite{bird}, and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

 Such results are in accordance with phenomena observed in this work. Linguistic characterization is being described in a deriving article~\cite{rcText}. See Appendix~\ref{sec:fure} for further considerations of related work.

\section{Data description}
    \subsection{Email lists and messages}
Email list messages were obtained from
the GMANE email archive~\cite{GMANE}, which consists of more than 20,000 email lists and more than 130,000,000 messages~\cite{GMANEwikipedia}. These lists cover a variety of topics, mostly technology-related. It can be described as a corpus with metadata of its messages, such as time, place, sender name, sender email address.
GMANE usage in scientific research is reported in studies of isolated lists and of lexical innovations~\cite{GMANE2,bird}. Appendix~\ref{scripts} is dedicated scripts for gathering and processing GMANE email messages.

    \subsection{Chosen dataset}
The four lists below were selected for the resulting diversity, easing initial observance of natural and general properties.
\begin{itemize}
    \item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}. Dominated by participants with hybrid artistic and technological interests. Participants are from different countries, and English is the language used the most. Abbreviated as LAU from now on.
    \item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}. Participants are from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
    \item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}. Dominated by specialized computer programmers. Participants are from different countries, and English is the language used the most. Abbreviated as CPP from now on.
    \item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}. Dominated by Brazilian activists and digital culture interests. Participants are mostly Brazilians, and Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

@@

\section{Characterization methods}
After immersion with the data and appropriated literature, the following methods for network characterization were chosen: 1) statistics of activity along time; 2) division of network in hubs, intermediary and peripheral vertex; 3) prominence of topological measures; 4) evolutive visualizations and quantitative observations; 5) typological elaborations of networks and participants.
Each of these methods are described bellow with supporting structures.

    \subsection{Temporal activity statistics}
Number of messages along time with respect to seconds, minutes, hours, days of the week, days of the month, and months of the year. These are exhibited as tables in Appendix~\ref{tabTime}. Results are outlined in Section~\ref{constDisc}.

    \subsection{Interaction network}\label{intNet}
Regarding literature~\cite{bird,newmanCommunityDirected,newmanCommunity2013}, interaction networks can be modeled both weighted or unweighted, both directed or undirected. Networks in this article are directed and weighted, considered as more informative among possibilities (directed unweighted, undirected weighted, and undirected unweighted).
More precisely, the networks reported are erected as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network, as B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to edge weight. Self-loops were regarded as non-informative and discarded. This networks are described as exhibiting free-scale and small world properties, as expected for a social network~\cite{bird}.

@@

Previous messages on the thread create directed edges from their author to the observed message's author. Edges can be created from all antecedent messages on the message-response thread. In this work, only the immediate predecessor are linked to new message's author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a connection between $x\rightarrow z$. Potential interpretations for this weaker connection are usually common sense, such as: double length, half weight or with one more ``obstacle''. This suggests the adoption of other centrality measures that account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

@@

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser}
    \caption{Degree distribution on scale-free and Erd\"os-R\'enyi ideal networks. The later has more
        intermediary, as the former has more peripheral and hub vertexes. Sections are
        given by the two intersections of the distributions $k_1$ and $k_2$. Characteristic degrees
    are in compact intervals of degree: $[0,k_1]$, $(k_1,k_2]$, $(k_2,k_{max}]$ for the three sections considered (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

Because of social networks tendency to have a scale-free distribution of connectivity,
one can compare it to an Erd\"os-R\'enyi random graph and consider peripheral, intermediary and hub
sectors~\cite{3setores}, as depicted in Figure~\ref{fig:setores}.
The degree distribution $\widetilde{P}(k)$ of an ideal
scale-free network $\mathcal{N}_f$ with $N$ vertexes and $z$ edges, has less
average degree nodes when compared with the distribution $P(k)$ of an Erd\"os-R\'enyi
random graph with the same number of vertexes and edges:

@@

If $\mathcal{N}_f$ is directed and has no self-loops, the probability
of the presence of an edge between two arbitrary vertexes is $p_e=\frac{z}{N(N-1)}$ (see Appendix~\ref{ap:ded}).
A vertex in the ideal Erd\"os-R\'enyi digraph with the same number of vertexes and edges, and thus the same probability $p_e$ for the presence of an edge, will have degree $k$ with probability:

@@

The arguments behind this classification are: 1) vertexes so connected that they are virtually inexistent in networks whose edges are incident regardless of vertex properties, are correctly associated to the hubs sector. Vertexes with very few connections, which are way more abundant than expected in networks whose edges are incident regardless of vertex properties, are correctly associated to periphery. Degree values near average, predicted as the most abundant if connections are created without criteria, and less frequent in free-scale phenomena, are correctly associated to intermediary vertexes.

To assure statistical validity, bins can be chosen to contain at least $\eta$ vertexes. Thus, each bin, starting at degree $k_i$, spans $\Delta_i=[k_{i},k_{j}]$ degree values, where $j$ is the smallest integer in which degrees $k_i$-$k_{j}$ contain at least $\eta$ values. This changes equation~\ref{criterio} to:

\begin{equation}\label{criterio2}
    \sum_{x=k_i}^{k_j} \widetilde{P}(x) < \sum_{x=k_i}^{k_j} P(x) \Rightarrow \text{i is intermediary}
\end{equation}

If instead strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}$ the average weight of an edge and $s_i$ the vertex strength. For in and out degrees and strengths, comparisons should be made with $\kappa_i=2k_i^{in}$, $\kappa_i=2k_i^{out}$, $\kappa_i=2\frac{s_i^{in}}{\overline{w}}$ and $\kappa_i=2\frac{s_i^{out}}{\overline{w}}$. Results of these criteria for network segmentation are discussed in subsection~\ref{subsec:pih}.

@@

\begin{itemize}
    \item Exclusivist criteria: vertex are only classified if the class is the same with respect to all measures. In this case, total vertex classified (usually) does not reach 100\%, which is depicted by a black line in Appendix~\ref{figures}.
    \item Inclusivist criteria: vertex has class given by any of the measures. In this case, a vertex can have more than one class and the fraction of class attribution beyond total number of vertexes is also depicted by a black line in Appendix~\ref{figures}.
    \item Exclusivist cascade: hubs are only hubs if classified as hub with respect to all measures. Intermediary are the vertexes classified as intermediary or hub with respect to all measures. Vertexes left are regarded as peripheral vertex.
    \item Inclusivist cascade: hubs are vertexes classified as hubs by any of the measures. From the vertexes left, if any is classified as intermediary by any measure, than it is intermediary. The rest of the vertexes are peripheral.
    \item Exclusivist externals: hubs have unanimous classification with respect to all measures. Of vertexes left, peripheral vertexes are the ones classified as hub or peripheral by simple criterion. The rest represent intermediary sector.
    \item Inclusivist externals: hubs are vertexes classified as hubs with respect to any measure. From vertexes left, if a vertex is classified as peripheral with respect to any measure, than it is peripheral. The rest is regarded intermediary sector.
\end{itemize}

@@

        \subsubsection{Topological measures}\label{measures}
This article
restricts topological analysis to a small selection of the most basic
measures of each vertex:

\begin{itemize}
    \item Degree $d_i$: number edges linked to node $i$.
    \item In-degree $d_i^{in}$: number of edges ending at node $i$.
    \item Out-degree $d_i^{out}$: number of edges departing from node $i$.
    \item Strength $s$: sum of weights of all edges linked to node $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at node $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from node $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked. This is the standard clustering coefficient for undirected graphs. Usage of other clustering coefficients (e.g. for directed graphs) was considering out of scope, as such clustering coefficients are usually employed in special contexts. 
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain the node $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

@@

\begin{itemize}
    \item asymmetry of vertex $i$: $asy_i=\frac{d_i^{in}-d_i^{out}}{d_i}$.
    \item mean of edge asymmetry of vertex $i$: $\mu_i^{asy}=\frac{\sum_{j\in J_i} e_{ji}-e_{ij}}{|J_i|}$. Where $e_{xy}$ is 1 if there is and edge from $x$ to $y$, $0$ otherwise. $|J_i|$ is the number of neighbors of vertex $i$.
    \item standard deviation of edge asymmetry of vertex $i$: $\sigma_i^{asy}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{asy} -(e_{ji}-e_{ij}) ]^2  }{|J_i|}  }$
    \item disequilibrium of vertex $i$: $dis_i=\frac{s_i^{in}-s_i^{out}}{s_i}$.
    \item mean of edge disequilibrium of vertex $i$: $\mu_i^{dis}=\frac{\sum_{j \in J_i}\frac{w_{ji}-w_{ij}}{s_i}}{|J_i|}$, where $w_{xy}$ is weight of edge $x\rightarrow y$ and zero if there is no such edge.
    \item standard deviation of edge disequilibrium of vertex $i$: $\sigma_i^{dis}=\sqrt{\frac{\sum_{j\in J_i}[\mu_{dis}-\frac{(w_{ji}-w_{ij})}{s_i}]^2}{|J_i|}}$
\end{itemize}

   \subsection{Evolutive observations}
Evolution of network is observed within a fixed number of messages (window size: $ws$) that shifts in the message timeline.
 All $ws=$ 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000 were used. Within a same $ws$, the number vertexes and edges vary in time, as do other network characteristics. Further work should deepen inspection of measure interdependence, this article holds to measures in Section~\ref{measures}.

        \subsubsection*{Visualization of network evolution}
In refining hypotheses, visualization of the network was crucial. Animations, image galleries and online gadgets were made~\cite{animacoes,galGMANE,appGMANE}. Mapping of various topological measures to glyphs and layouts are being further explored as a parallel research. Furthermore, stable aspects of measures prominence along time are captured through mean and standard deviation (see Section~\ref{measures} and Appendix~\ref{sec:pcat}). Constant sector sizes along time are observed in a timeline fashion in Appendix~\ref{figures}.

@@

        \subsubsection{Hours of the day}
Table~\ref{dia} shows how the four lists distribute activity along the day.
Afternoon was the most active 6h period of the day. Second 12h more active than first 12h. Even so, activity peak occurs around midday, with a skew towards earlier hours.

        \subsubsection{Days of the week}
 Weekdays also exhibit an interesting pattern, to which is dedicated Table~\ref{semana}: a decrease of activity on weekends of at least one third, reaching two thirds in extreme cases.
        \subsubsection{Days along the month}
 Table~\ref{mes} shows activity along the month, in which no prevalent pattern was observed.
 Variation of activity in the days along the month is less prominent, one cannot point much more than a - probably not statistically relevant - tendency of first and second weeks to be more active. The most important trait might be their homogeneity with respect to activity. Last days of the month (29, 30 and 31) are not present in every month, and observed activity is proportional to incidence rates.

        \subsubsection{Months and larger divisions of the year}
Table~\ref{ano} is dedicated to activity in months and larger divisions of the year.
Observation points two periods of more prominent activities: Jun-Aug (MET and LAD), and Dec-Mar (CPP, LAU and LAD), which fit academic calendars, vacations and end-of-year holidays.

@@

As specified in Section~\ref{sectioning}, in order to classify nodes as hubs, intermediary or peripheral, the incident connectivity distribution $\widetilde{P}(k)$ is compared
to a connectivity distribution of an ideal Erd\"os R\`nyi $P(k)$ with the same number of vertexes and edges. In the networks inspected for this article (see Section~\ref{intNet}), if degree distribution is used for classification, hubs sector size reaches peaks with $10\%$ of all vertexes. 
 If strength is used for comparison ($\widetilde{P}(k_i=\frac{s_i}{\overline{w}})$) is compared
to $P(k_i)$), hubs account for approximately $5\%$ of all vertex, i.e. strength classification yields half the number of hubs as plain degree. 
This results hold for in and out degrees and strengths,

Classification criteria exposed in Section~\ref{sectioning} was used efficiently with windows of at least 200 messages. Specially with 1000 or more messages, criteria yields stable fractions of $\approx 5\%$ of hubs, $\approx [15-20]\%$ of intermediary and $\approx [75-80]\%$ peripheral vertex, which match literature~\cite{priSect}. A compound criteria, also described in Section~\ref{sectioning}, can be used as a classification refinement. This is specially useful in dealing with fewer messages, in which case the structure degenerates with respect to some of the degree and strength measures, but not all.
A minimum window size for observation of more general properties can be inferred by monitoring the giant component and the degeneration of hub, intermediary and peripheral sections. This degeneration is critical in the span of 50-100 messages for window sizes. With compound criteria, such as exclusive cascade of Figure~\ref{fig:cpp250_}, the networks seem to hold basic structure even with as few as 20-50 messages. This indicates that concentration of activity and low-activity participants occurs even with very few messages.

For the histograms used in the classification process, the usage of at least $\eta$ vertexes for each bin did not yield significant differences.
That was understood as a consequence of the observation scale:
\emph{There are between 20 and 1200 participants in the message window sizes used to derive most results ($ws \in [200,1500]$ messages). As peripheral vertexes are abundant and span few degrees, there are more than $\eta$ vertexes with each low degree value. Higher degrees cases requires a bit more reasoning. Specified $ws$ implicates that each participant is $p \in [0.1\%,0.5\%]$ of all participants. Therefore, if incident connectivity is very improbable in an Ed\"os R\`enyi network (less than $p$, the probability that a single participant represent when histogram is normalized to density function), than it is not an intermediary connectivity, but a hub.} This concludes the explanation of why using at least $\eta$ vertexes for each bin did not impact results.

@@

This work aimed at finding common characteristics among (email) interaction networks. Analysis involved primary measures observance and formal criteria for coherent ratios of hub, intermediary and hub sectors. Nevertheless, inspection done by visualizations and raw data manipulations suggests agents peculiarities of typological character. These are initial observations, which should be further developed as posed in Section~\ref{sec:ftyp}:
